---
title: "SATC Analysis"
author: "R. Axel W. Wiberg"
date: "31.01.2022" # Last updated
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Source the SATC functions
```{r}
source(here::here("scripts/SATC/","satcFunc.R"))
library(reshape)
library(ggplot2)
```


Load indxstats files
```{r}
files<-list.files(here::here("SATC/filt_readsreps"),pattern = ".idxstats")
idx<-lapply(paste(here::here(),"/SATC/filt_readsreps/",files,sep=""), read.table,as.is=TRUE)
names(idx)<-gsub(".idxstats","",files)
names(idx)<-gsub("_filtrepsreads","",names(idx))

cov_mat<-sapply(idx,function(x) x$V3)
rownames(cov_mat)<-idx$CmacF1$V1

covlen_mat<-sapply(idx,function(x) x$V3/x$V2)
rownames(covlen_mat)<-idx$CmacF1$V1
covlen_mat_df<-as.data.frame(covlen_mat)
covlen_mat_df$scaf<-row.names(covlen_mat)
covlen_mat_df$len<-idx$CmacF1$V2
covlen_mat_df_m<-melt(covlen_mat_df,id.vars = c("scaf","len"))
colnames(covlen_mat_df_m)<-c("scaf","len","sample","covlen")

ggplot()+
  geom_point(data=covlen_mat_df_m,aes(x=log(len),y=covlen,fill=sample),shape=21,alpha=1/3)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  xlab("log(length)")+
  ylab("Nreads/Length")#+
```


List of Philipp's candidate Y scaffolds
```{r}
philY<-c("utg000322l_1","utg000312c_1","utg000610l_1", "utg000584l_1", "utg001235l_1")
philX<-c("utg000057l_1", "utg000359l_1","utg000114l_1", "utg000326l_1", "utg000139l_1",
         "utg000191l_1", "utg000532l_1")
```


Filter data and compute normalised coverage
```{r}
## Filter scafoolds (min 100kb ) and normalize using the M longest scaffold 
rFilt <- filterScaffold(dat=idx,minLength=1e5,M=5,range=c(0,2))
```


Plot coverage
```{r}
covlen_mat<-sapply(rFilt,function(x) x$Nreads/x$Length)
rownames(covlen_mat)<-rFilt$CmacF1$scaffold
covlen_mat_df<-as.data.frame(covlen_mat)
covlen_mat_df$scaf<-row.names(covlen_mat)
covlen_mat_df$len<-rFilt$CmacF1$Length
covlen_mat_df$normscaf<-rFilt$CmacF1$normScafs
covlen_mat_df_m<-melt(covlen_mat_df,id.vars = c("scaf","len","normscaf"))
colnames(covlen_mat_df_m)<-c("scaf","len","normscaf","sample","covlen")

ggplot()+
  geom_point(data=covlen_mat_df_m,aes(x=log(len),y=covlen,fill=sample),shape=21,alpha=1/3)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  geom_point(data=covlen_mat_df_m[covlen_mat_df_m$normscaf,],
             aes(x=log(len),y=covlen,group=sample),shape=21,size=5,colour="black")+
  xlab("log(length)")+
  ylab("Nreads/Length")

cov_mat <- sapply(rFilt,function(x) x$norm)
cov_mat_df<-as.data.frame(cov_mat)
cov_mat_df$scaf<-rFilt$CmacF1$scaffold
cov_mat_df$len<-rFilt$CmacF1$Length
cov_mat_df$normscaf<-rFilt$CmacF1$normScafs
cov_mat_df_m<-melt(cov_mat_df,id.vars = c("scaf","len","normscaf"))
colnames(cov_mat_df_m)<-c("scaf","len","normscaf","sample","normcov")
head(cov_mat_df_m)

cov_mat_df_m[cov_mat_df_m$normscaf,]
ggplot()+
  geom_point(data=cov_mat_df_m,aes(x=log(len),y=normcov,fill=sample),shape=21,alpha=1/3,size=2)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  geom_point(data=cov_mat_df_m[cov_mat_df_m$normscaf,],
             aes(x=log(len),y=normcov,group=sample),shape=21,size=5,colour="black")+
  xlab("log(length)")+
  ylab("Normalised coverage")

```


The first scaffold is longest and has much higher coverage than the others. Try normalising with just the next four.
Try not using: "utg000012l_1", "utg000020l_1"
```{r}
normScafs<-c("utg000123l_1","utg000010l_1","utg000072l_1")
## Filter scafoolds (min 100kb ) and normalize using the M longest scaffold 
rFilt <- filterScaffold(dat=idx,minLength=1e5,normScaffolds = normScafs,range=c(0,2))
```

Plot coverage
```{r}
covlen_mat <- sapply(rFilt,function(x) x$Nreads/1000000)
rownames(covlen_mat)<-rFilt$CmacF1$scaffold

covlen_mat_df<-as.data.frame(covlen_mat)
covlen_mat_df$scaf<-row.names(covlen_mat)
covlen_mat_df$len<-rFilt$CmacF1$Length
covlen_mat_df$normscaf<-rFilt$CmacF1$normScafs
covlen_mat_df_m<-melt(covlen_mat_df,id.vars = c("scaf","len","normscaf"))
colnames(covlen_mat_df_m)<-c("scaf","len","normscaf","sample","covlen")

covlen_mat_df_m[covlen_mat_df_m$normscaf,]

ggplot()+
  geom_point(data=covlen_mat_df_m,aes(x=log(len),y=covlen,fill=sample),shape=21,alpha=1/3)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  geom_point(data=covlen_mat_df_m[covlen_mat_df_m$normscaf,],
             aes(x=log(len),y=covlen,group=sample),shape=21,size=5,colour="black")+
  xlab("log(length)")+
  ylab("Nreads (x 1,000,000)")#+

cov_mat <- sapply(rFilt,function(x) x$norm)
cov_mat_df<-as.data.frame(cov_mat)
cov_mat_df$scaf<-rFilt$CmacF1$scaffold
cov_mat_df$len<-rFilt$CmacF1$Length
cov_mat_df$normscaf<-rFilt$CmacF1$normScafs
cov_mat_df_m<-melt(cov_mat_df,id.vars = c("scaf","len","normscaf"))
colnames(cov_mat_df_m)<-c("scaf","len","normscaf","sample","normcov")

cov_mat_df_m[cov_mat_df_m$normscaf,]
ggplot()+
  geom_point(data=cov_mat_df_m,aes(x=log(len),y=normcov,fill=sample),shape=21,alpha=1/3,size=2)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  geom_point(data=cov_mat_df_m[cov_mat_df_m$normscaf,],
             aes(x=log(len),y=normcov,group=sample),shape=21,size=5,colour="black")+
  xlab("log(length)")+
  ylab("Normalised coverage")#+
```


```{r}
##plot normalized depth
plotDepth(rFilt) ##all scaffs
plotDepth(rFilt,normOnly=TRUE) ##scaffolds used for normalization (look for outliers)

## identify sex and sex scaffolds
sex <- sexDetermine(dat=rFilt, K=2, model="gaussian")
str(sex)
## See the inferred sex of each indiviual
head(cbind(names(idx),sex$sex))

## plot the clustering
mat <- sapply(sex$dat,function(y) y$norm)
sex_pca_dat<-as.data.frame(sex$pca$x)
sex_pca_dat$sample<-colnames(mat)
sex_pca_dat$sex<-sex$sex
head(sex_pca_dat)

ggplot()+
  geom_point(data=sex_pca_dat,aes(x=V1,y=V2,fill=sex),shape=21,size=3)+
  geom_label(data=sex_pca_dat,aes(x=V1,y=V2,label=sample),vjust=1.5)+
  scale_x_continuous("PC1",limits=c(-2,2))+
  scale_y_continuous("PC2",limits=c(-0.6,0.6))+
  scale_fill_discrete("Inferred sex",labels=c("heteromorphic\n(XY)","homomorphic\n(XX)"))+
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=16))

plotGroup(sex)

## See the inferred status of each scaffold
head(sex$SexScaffolds)
sex$SexScaffolds[match(c(philY), sex$SexScaffolds$Name),]
sex$SexScaffolds[match(c(philX), sex$SexScaffolds$Name),]

sig_scafs<-sex$SexScaffolds[which(sex$SexScaffolds$Pval<0.05),]


length(which(sex$SexScaffolds$Name[which(sex$SexScaffolds$Pval<0.05)] %in% c(philX,philY)))
nrow(sex$SexScaffolds[which(sex$SexScaffolds$Pval<0.05),])
table(sex$SexScaffolds$betaval[which(sex$SexScaffolds$Pval<0.05)] > 0)
```

Plot the data for Phil's contigs
```{r}
mat <- sapply(sex$dat,function(y) y$norm)
rownames(mat) <- as.character(sex$dat[[1]][,1])

phil_scafs<-mat[c(philX,philY),]
phil_scafs<-t(phil_scafs)
phil_scafs<-as.data.frame(phil_scafs)
phil_scafs$samp<-row.names(phil_scafs)
phil_scafs$sex<-sex$sex

phil_scafs_m<-melt(phil_scafs,id.vars = c("samp","sex"))
colnames(phil_scafs_m)<-c("samp","sex","contig","norm")
ggplot()+
  geom_point(data=phil_scafs_m,aes(x=contig,y=norm,fill=sex,group=sex),
             shape=21,size=3,position=position_dodge(0.2),alpha=1/2)+
  ylab("Normalised Coverage")+
  xlab("")+
  theme(axis.text.x = element_text(angle=45,vjust=1,hjust=1))
```

Plot all scafs with significantly different coverage 
+ scafs that are labelled by Phil
```{r}
mat <- sapply(sex$dat,function(y) y$norm)
rownames(mat) <- as.character(sex$dat[[1]][,1])

sig_diff_scafs<-sex$SexScaffolds$Name[which(sex$SexScaffolds$Pval<0.05)]
sig_dif_scafs$type<-rep("Unnkown",nrow(sig_dif_scafs))
sig_dif_scafs$type[match(c(philX), sig_dif_scafs$Name)]<-"X"
sig_dif_scafs$type[match(c(philY), sig_dif_scafs$Name)]<-"Y"

sig_dif_scafs_dat<-mat[unique(c(philX,philY,sig_diff_scafs)),]
sig_dif_scafs_dat<-t(sig_dif_scafs_dat)
sig_dif_scafs_dat<-as.data.frame(sig_dif_scafs_dat)
sig_dif_scafs_dat$samp<-row.names(sig_dif_scafs_dat)
sig_dif_scafs_dat$sex<-sex$sex
sig_dif_scafs_dat<-melt(sig_dif_scafs_dat,id.vars = c("samp","sex"))
colnames(sig_dif_scafs_dat)<-c("samp","sex","contig","norm")

ggplot()+
  geom_point(data=sig_dif_scafs_dat,
             aes(x=contig,y=norm,fill=sex,group=sex),
             shape=21,size=3,position=position_dodge(0.2),alpha=1/2)+
  ylab("Normalised Coverage")+
  xlab("")+
  scale_fill_discrete("Inferred sex",labels=c("heteromorphic\n(XY)","homomorphic\n(XX)"))+
  theme(axis.text.x = element_text(angle=45,vjust=1,hjust=1))

# Only plot scaffolds with abs(beta) > 0.1
beta0.1<-sex$SexScaffolds$Name[which(abs(sex$SexScaffolds$betaval)>0.1)]

# If beta is negative, then homomorphic coverage is lower than heteromorphic
putativeY<-sex$SexScaffolds$Name[which(sex$SexScaffolds$betaval<0)]
putativeY<-putativeY[which(putativeY%in%as.character(sig_dif_scafs_dat$contig))]
# If beta is negative, then homomorphic coverage is lower than heteromorphic
putativeX<-sex$SexScaffolds$Name[which(sex$SexScaffolds$betaval>0)]
putativeX<-putativeX[which(putativeX%in%as.character(sig_dif_scafs_dat$contig))]
sig_dif_scafs_dat$contig<-factor(sig_dif_scafs_dat$contig,levels=c(putativeX,putativeY))

sig_dif_scafs_dat$type<-rep("Unknown",nrow(sig_dif_scafs_dat))
sig_dif_scafs_dat$type[which(sig_dif_scafs_dat$contig %in% c(philX))]<-"(Phil)Putative X"
sig_dif_scafs_dat$type[which(sig_dif_scafs_dat$contig %in% c(philY))]<-"(Phil)Putative Y"
sig_dif_scafs_dat

#[sig_dif_scafs_dat$contig %in% beta0.1,]

ggplot()+
  geom_point(data=sig_dif_scafs_dat,
             aes(x=contig,y=norm,fill=sex,group=sex,size=type),
             shape=21,position=position_dodge(0.2),alpha=1/2)+
  ylab("Normalised Coverage")+
  xlab("")+
  scale_fill_discrete("Inferred sex",labels=c("heteromorphic\n(XY)","homomorphic\n(XX)"))+
  scale_size_manual("",values=c(5,5,2),guide=FALSE)+
  theme(axis.text.x = element_text(angle=45,vjust=1,hjust=1),
        axis.text=element_text(size=16),
        axis.title=element_text(size=16),
        legend.text=element_text(size=16),
        legend.title=element_text(size=16))


sex$SexScaffolds$linkage<-ifelse(sex$SexScaffolds$betaval>0,"X","Y")
sex$SexScaffolds$other_ev<-rep("0",nrow(sex$SexScaffolds))
sex$SexScaffolds$other_ev[which(sex$SexScaffolds$Name %in% c(philX))]<-"1"
sex$SexScaffolds$other_ev[which(sex$SexScaffolds$Name %in% c(philY))]<-"1"

sex$SexScaffolds[which(sex$SexScaffolds$Name %in% unique(c(philX,philY,sig_diff_scafs))),]
nrow(sex$SexScaffolds[which(sex$SexScaffolds$Name %in% unique(c(philX,philY,sig_diff_scafs))),])
```


```{r}
# What is coverage at the contigs used for normalising
cov_mat[rFilt$CmacF1$scaffold[which(rFilt$CmacF1$normScafs)],]
# What is coverage at phil candidate Y
cov_mat[philY]
# What is coverage at scaffolds with significantly different coverage
cov_mat[c(putativeX,putativeY),]
```


```{r}
mat_first_df<-as.data.frame(mat_first)
mat_first_df$scaf<-row.names(mat_first)
mat_first_df$len<-rFilt$CmacF1$Length
mat_first_df$normscaf<-rFilt$CmacF1$normScafs
mat_first_df_m<-melt(mat_first_df,id.vars = c("scaf","len","normscaf"))
colnames(mat_first_df_m)<-c("scaf","len","normscaf","sample","normcov")
head(mat_first_df_m)

mat_first_df_m[mat_first_df_m$normscaf,]
ggplot()+
  geom_point(data=mat_first_df_m,aes(x=log(len),y=normcov,fill=sample),shape=21,alpha=1/3,size=2)+
  scale_fill_manual("Sample",values=c("pink","red","lightblue","blue"))+
  geom_point(data=mat_first_df_m[mat_first_df_m$normscaf,],
             aes(x=log(len),y=normcov,group=sample),shape=21,size=5,colour="black")+
  geom_point(data=mat_first_df_m[which(mat_first_df_m$scaf%in%c(putativeX,putativeY)),],
             aes(x=log(len),y=normcov),shape=21,size=5,colour="red")+
  xlab("log(length)")+
  ylab("Normalised coverage")#+

```


Plot 20 random scaffolds not in Phil's list or with different coverage
```{r}
names(sex$dat)

mat <- sapply(sex$dat,function(y) y$norm)
rownames(mat) <- as.character(sex$dat[[1]][,1])
head(mat)
head(t(mat))

todrop<-unique(c(philY,philX,sig_diff_scafs))
remaining_scafs<-sex$SexScaffolds$Name[which(!(sex$SexScaffolds$Name %in% todrop))]
random_scafs<-sample(remaining_scafs,size = 30,replace = FALSE)
random_scafs_dat<-mat[random_scafs,]
random_scafs_dat<-t(random_scafs_dat)
random_scafs_dat<-as.data.frame(random_scafs_dat)
random_scafs_dat$samp<-row.names(random_scafs_dat)
random_scafs_dat$sex<-sex$sex

random_scafs_dat_m<-melt(random_scafs_dat,id.vars = c("samp","sex"))
colnames(random_scafs_dat_m)<-c("samp","sex","contig","norm")

sex$SexScaffolds[which(sex$SexScaffolds$Name %in% remaining_scafs),]
head(random_scafs_dat_m)

ggplot()+
  geom_point(data=random_scafs_dat_m,aes(x=contig,y=norm,fill=sex,group=sex),
             shape=21,size=3,position=position_dodge(0.2),alpha=1/2)+
  ylab("Normalised Coverage")+
  xlab("")+
  theme(axis.text.x = element_text(angle=45,vjust=1,hjust=1))
```

